---
title: "Severe Weather Events in The USA"
author: "Mateus Melo"
date: "05/08/2020"
output:
        html_document:
                 keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract


## Loading and Processing The Data

We start our analysis by downloading  and reading the [storm data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2). Then we check the number of variables and observations.

```{r cache=TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","data.bz2")
df<-read.csv("data.bz2")
dim(df)
```

We see that the dataset is very large, having 37 variables and 902297 observations. Since the focus of our analysis are the population health and economic impacts, we can remove many of the variables, improving the performance of the code. Then, we can check the type and format of the data.

```{r cache=TRUE}
df<-subset(df, select = c("BGN_DATE", "STATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP"))
str(df)
```

We are left with 9 variables now. The first  being  the BGN_DATE, which corresponds to the event date. It is useful to change the variable date type. To do so, we are going to use the lubridate and stringr packages.

```{r cache=TRUE, message=FALSE, warning=FALSE}
library(lubridate)
library(stringr)
df$BGN_DATE<- mdy(str_split(df$BGN_DATE," ", simplify = TRUE) [,1])
```

Now we are going to take a deeper look into the variables.

```{r}
summary(df)
```

The events dates range from 1950 to 2011, a 61 years interval, but 50% of the data was recorded after 1998 and 25% was recorded after 2007. Being so, we are going to focus our analysis in the latter period, since our social/economic/technology context is more related to this period and doing that, we also improve the code performance and time consuming. Another thing we can do to improve the performance is to keep track only of the events that have had some kind of impact in the population health or economy. We can see in the summary above that most of the events have not caused any deaths (FATALITIES variable), injuries (INJURY variable), property damage (PROPDMG variable) or crop damage (CROPDMG).

```{r cache=TRUE}
df<-subset(df, BGN_DATE >= ymd("2007-07-28") & (FATALITIES > 0 | INJURIES > 0 | PROPDMG > 0 | CROPDMG >0))
dim(df)
```

Now we are left with 72869 observations.


## Cleaning The Data

In the summary above is shown that the crop and property damage are broken in two columns for each, respectively, CROPDM, CROPDMEXP, PROPDM and PROPDMEXP. That breaks one of the tidy data principles and must be corrected. First, we are going to check the PROPDMGEXP and CROPDMGEXP. 
```{r}
table(df$PROPDMGEXP)
table(df$CROPDMGEXP)
```

We here assume that K, M and B represent respectively 10<sup>3</sup>, 10<sup>6</sup> and 10<sup>9</sup> factors. Let's put it all together in a single column for each variable.

```{r cache=TRUE}
df$CROPAUX <- ifelse(df$CROPDMGEXP=="K",3,ifelse(df$CROPDMGEXP=="M",6,9))
df$PROPAUX <- ifelse(df$PROPDMGEXP=="K",3,ifelse(df$PROPDMGEXP=="M",6,9))
df$CROPDMG <- df$CROPDMG * 10 ^ df$CROPAUX
df$PROPDMG <- df$PROPDMG * 10 ^ df$PROPAUX
df <- df<-subset(df, select = c("BGN_DATE", "STATE","EVTYPE","FATALITIES","INJURIES","PROPDMG","CROPDMG"))
```

Now we are going to take a look in the event type variable, which is corresponded by the EVTYPE column. The event types must be as described in the [storm data documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

```{r}
table(df$EVTYPE)
```

All of the events agree with the documentation list with the exception of the *LANDSLIDE* type. That difference is also explained in  the documentation and the description should be replaced by the *DEBRIS FLOW*.

```{r}
df$EVTYPE[df$EVTYPE=="LANDSLIDE"] <- "DEBRIS FLOW"
```

Now the data is tidy and we are ready to start the analysis.